Binding of cations to phospholipid bilayers

### 2015-02-18 LIPIDionINTERACT/LIPIDionINTERACT.tex
fix the ion concetrations

\caption{Simulated lipid bilayers with ions.
}\label{IONsystems}

@@

Berger-POPC-07\cite{ollila07a}, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}   &   POPC & 190 (NaCl) & 128 & 7202 & 44  & 0  & 44 &298  & 110 & 50 &?\todoi{Samuli put to Zenodo} \\
Berger-POPC-07\cite{ollila07a}, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}   &   POPC & 190 (CaCl$_2$) & 128 & 7157 & 0 & 44  & 88 &298 & 110 & 50 &?\todoi{Samuli put to Zenodo}  \\

@@

CHARMM36\cite{klauda10}, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}   & POPC & 140 (NaCl)  & 72 & 2085 & 13  & 0 & 13 & 303  & 80 & 60 &?\todoi{Samuli put to Zenodo}\\
CHARMM36\cite{klauda10}, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}   & POPC & 280 (NaCl)  & 72 & 2085 & 26  & 0 & 26 & 303  & 80 & 60 &?\todoi{Samuli put to Zenodo} \\
CHARMM36\cite{klauda10}, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}  & POPC & 570 (NaCl)  & 72 & 2085 & 52  & 0 & 52 & 303  & 80 & 60 &?\todoi{Samuli put to Zenodo} \\
CHARMM36\cite{klauda10}, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}   & POPC & 85  (NaCl)  & 72 & 2085 & 78  & 0 & 78 & 303  & 73 & 60 &?\todoi{Samuli put to Zenodo} \\

@@

MacRog\cite{maciejewski14}\todoi{DONE} , ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}        & POPC &  200 (NaCl) & 288 & 14500 & 54 & 0 & 54 & 310 & 90&50  &\cite{macrogIONfiles}  \\
MacRog\cite{maciejewski14}\todoi{DONE} , ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}        & POPC &   300 (NaCl) & 288 & 14446 & 81 & 0 & 81 & 310 & 90&50  & \cite{macrogIONfiles} \\
MacRog\cite{maciejewski14}\todoi{DONE} , ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}        & POPC &   400 (NaCl) & 288 & 14392 & 108 & 0 & 108 & 310 & 90& 50  & \cite{macrogIONfiles}  \\

@@

Orange, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}   &   POPC & 250 (NaCl) & 72 & 2802 & 26 & 0  & 26 & 298 & 120 & 60 &?\todoi{Jukka Määttä and Luca Monticelli, please let us know if we can share some files through Zenodo. This is unpublished model.}  \\
Orange, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}   &   POPC & 240 (CaCl$_2$)  & 72 & 2802 & 0 & 26  & 52 & 298 & 120 & 60 &? \todoi{Jukka Määttä and Luca Monticelli, please let us know if we can share some files. This is unpublished model.} \\

@@

%The experimental results~\cite{akutsu81}, plotted in Fig.~\ref{ordPions} shows a distinct difference
%in behaviour as a function of ions: the choline order parameters were unaffected by the addition of NaCl, 
%whereas a systematic decrease was observed by the addition of CaCl$_2$. Following the electrometer
%concept this means that the binding constant for divalent Ca$^{2+}$ is higher than for monovalent Na$^+$
%which is in line with the large amount of experimental data with different techniques~\cite{cevc90,binder02,petrache06,pabst07,filippov09}.
\begin{figure}[]
  \centering
  \includegraphics[width=8cm]{OrderParameterIONSchanges.eps}
  \caption{\label{ordPions}
    The change of order parameters for $\beta$ and $\alpha$ segments as a function of NaCl (left column) and CaCl$_2$ (right column) concentrations from simulations 
    and experiments by Akutsu et al.~\cite{akutsu81}. The signs of the order parameters without ions are taken from~\cite{hong95a,hong95b,gross97} and it is reasonable
    to assume that the sign do not change with concentrations repsented here~\cite{altenbach84}. Note that the $\beta$ order parameter with negative sign decreases here with cation
    penetration in contrast to measured absolute value increase in the origical studues~\cite{akutsu81,altenbach84}.
    Regarding the actual electrometer concept this not relevant, however. 
    The experimental accuracy of order parameter changes is much higher than the accuracy of quantitative values due to high resolution of $^2$H NMR spectroscopy~\cite{??}. 
   }
\todo{I think that we have to run Orange simulation with higher NaCl concentration. I will do it.} \\
\todo{I think that we have to run CHARMM36 with CaCl$_2$. If anyone happen to have the data and is willing to share,
let us know. If not, then I will do it.} \\
\todo{ I think that it would be very interesting to test the modified CHARMM ions: Venable et al. dx.doi.org/10.1021/jp401512z, J. Phys. Chem. B 2013, 117, 10183−10192.
According to the paper, these parameters improve the ion binding to the charged lipid bilayers. Even though I think that testing this would be highly relevant, I do not have
time to do it now.}
\end{figure}

@@

To study the correlation between order parameter changes and ion partitioning into a bilayer, 
suggested by the electrometer concept~\cite{akutsu81,altenbach84,seelig87,scherer89}, the density distributions as a function of membrane normal 
are shown in Fig.~\ref{IONdensCOMP} for lipids and cations from different simulation models. 
%to see if larger ion induced effects
%in choline order parameters correlates with ion binding also in simulation models.
\begin{figure}[]
  \centering
  \includegraphics[width=8cm]{IONdensities.eps}
  \caption{\label{IONdensCOMP}
    Mass density profiles for lipids and ions from simulations with different force fields. 
    The lipid densities are divided with 100 to make them visible with the used y-axis scale.
  }
\todo{This figure is in progress, however, the main point can be seen. \\
- I think that we should use number densities instead of mass densities to make the difference between Na and Ca distributions more clear. 
Now the difference comparison is little bit distracted due to the different mass of the atoms. \\
- The lipid density from MacRog simulation is missing. Maybe Matti Javanainen could provide this?
If the data will be uploaded to Zenodo, it also trivial to calculate it with script. \\
- It seems that relative integrated (total) ion density in different models do not correspond the reported relative concentrations.
For example, total Na density seems bigger in Berger than in MacRog even though there is only minot difference in concentration, and into the wrong direction. \\
- I think that CHARMM36 simulation with CaCl$_2$ should be done \\
- Figure editing will be finished (centering of the graphs, labels, etc.) will be done once the above issues have been sorted out.
 }
\end{figure}
It is evident from the figure that the Na$^+$ density peaks at the lipid bilayer interface are higher (indicating stronger ion binding) for the models
experiencing also larger changes in order parameters with NaCl concentration shown in Fig.~\ref{ordPnacl}. 
The both Na$^+$ binding affinity and order parameter changes are smallest, thus closest to the experiments, for the Orange model, and then incease 
in order CHARMM36, MacRog, and Berger having the highest binding affinity and order parameter changes.
%Simultaneously, it is evident from Figs.~\ref{ordPnacl} and~\ref{ordPions} that the order parameter decrease due to addition
%of NaCl is more pronounced in Berger and MacRog models compared to the CHARMM36 and Orange. 
Thus, the correlation between decreasing choline order parameters
and amount of penetrating cations (i.e. the electrometer concept) is also seen in simulations, independently of the quantitative quality of the used simulation model.

>>>>

\caption{Simulated lipid bilayers with ions. The ion concentrations are the concentration of ions in buffer to solute the lipid bilayers and calculated as [ion]=(N$_{\rm ion} \times$[water])/N$_{\rm w}$, 
where [water]=55.5M. These correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.}\label{IONsystems}

@@

Berger-POPC-07\cite{ollila07a}, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}   &   POPC & 340 (NaCl) & 128 & 7202 & 44  & 0  & 44 &298  & 110 & 50 &?\todoi{Samuli put to Zenodo} \\
Berger-POPC-07\cite{ollila07a}, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}   &   POPC & 340 (CaCl$_2$) & 128 & 7157 & 0 & 44  & 88 &298 & 110 & 50 &?\todoi{Samuli put to Zenodo}  \\

@@

CHARMM36\cite{klauda10}, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}   & POPC & 350 (NaCl)  & 72 & 2085 & 13  & 0 & 13 & 303  & 80 & 60 &?\todoi{Samuli put to Zenodo}\\
CHARMM36\cite{klauda10}, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}   & POPC & 690 (NaCl)  & 72 & 2085 & 26  & 0 & 26 & 303  & 80 & 60 &?\todoi{Samuli put to Zenodo} \\
CHARMM36\cite{klauda10}, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}  & POPC & 1380 (NaCl)  & 72 & 2085 & 52  & 0 & 52 & 303  & 80 & 60 &?\todoi{Samuli put to Zenodo} \\
CHARMM36\cite{klauda10}, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}   & POPC & 2080  (NaCl)  & 72 & 2085 & 78  & 0 & 78 & 303  & 73 & 60 &?\todoi{Samuli put to Zenodo} \\

@@

MacRog\cite{maciejewski14}\todoi{DONE} , ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}        & POPC &  210 (NaCl) & 288 & 14500 & 54 & 0 & 54 & 310 & 90&50  &\cite{macrogIONfiles}  \\
MacRog\cite{maciejewski14}\todoi{DONE} , ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}        & POPC &   310 (NaCl) & 288 & 14446 & 81 & 0 & 81 & 310 & 90&50  & \cite{macrogIONfiles} \\
MacRog\cite{maciejewski14}\todoi{DONE} , ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}        & POPC &   420 (NaCl) & 288 & 14392 & 108 & 0 & 108 & 310 & 90& 50  & \cite{macrogIONfiles}  \\

@@

Orange, ionFF~\cite{??}\ &   POPC & 140 (NaCl) & 72 & 2866 & 7 & 0  & 7 & 298 & running & ? &?  \\
Orange, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}   &   POPC & 510 (NaCl) & 72 & 2802 & 26 & 0  & 26 & 298 & 120 & 60 &?\todoi{Jukka Määttä and Luca Monticelli, please let us know if we can share some files through Zenodo. This is unpublished model.}  \\
Orange, ionFF~\cite{??}  &   POPC & 1000 (NaCl) & 72 & 2780 & 50 & 0  & 50 & 298 & running & ? &? \\
Orange, ionFF~\cite{??}\todoi{Appropriate reference for the ion model?}   &   POPC & 510 (CaCl$_2$)  & 72 & 2802 & 0 & 26  & 52 & 298 & 120 & 60 &? \todoi{Jukka Määttä and Luca Monticelli, please let us know if we can share some files. This is unpublished model.} \\

@@

%The experimental results~\cite{akutsu81}, plotted in Fig.~\ref{ordPions} shows a distinct difference
%in behaviour as a function of ions: the choline order parameters were unaffected by the addition of NaCl, 
%whereas a systematic decrease was observed by the addition of CaCl$_2$. Following the electrometer
%concept this means that the binding constant for divalent Ca$^{2+}$ is higher than for monovalent Na$^+$
%which is in line with the large amount of experimental data with different techniques~\cite{cevc90,binder02,petrache06,pabst07,filippov09}.
\begin{figure}[]
  \centering
  \includegraphics[width=8cm]{OrderParameterIONSchanges.eps}
  \caption{\label{ordPions}
    The change of order parameters for $\beta$ and $\alpha$ segments as a function of NaCl (left column) and CaCl$_2$ (right column) concentrations from simulations 
    and experiments by Akutsu et al.~\cite{akutsu81}. The signs of the order parameters without ions are taken from~\cite{hong95a,hong95b,gross97} and it is reasonable
    to assume that the sign do not change with concentrations repsented here~\cite{altenbach84}. Note that the $\beta$ order parameter with negative sign decreases here with cation
    penetration in contrast to measured absolute value increase in the origical studues~\cite{akutsu81,altenbach84}.
    Regarding the actual electrometer concept this not relevant, however. 
    The experimental accuracy of order parameter changes is much higher than the accuracy of quantitative values due to high resolution of $^2$H NMR spectroscopy~\cite{??}. 
   }
\todo{I have already Orange simulations running with 1M and 140mM NaCl concentrations. I think I should still run CHARMM36 with 1M and maybe with close to 150mM. 
Some of the simulations had accidentally over 1M concentration and those are not useful for this plot.} \\
\todo{I think that we have to run CHARMM36 with CaCl$_2$. If anyone happen to have the data and is willing to share,
let us know. If not, then I will do it.} \\
\todo{ I think that it would be very interesting to test the modified CHARMM ions: Venable et al. dx.doi.org/10.1021/jp401512z, J. Phys. Chem. B 2013, 117, 10183−10192.
According to the paper, these parameters improve the ion binding to the charged lipid bilayers. Even though I think that testing this would be highly relevant, I do not have
time to do it now.}
\end{figure}

@@

To study the correlation between order parameter changes and ion partitioning into a bilayer, 
suggested by the electrometer concept~\cite{akutsu81,altenbach84,seelig87,scherer89}, the density distributions as a function of membrane normal 
are shown in Fig.~\ref{IONdensCOMP} for lipids and cations from different simulation models. 
%to see if larger ion induced effects
%in choline order parameters correlates with ion binding also in simulation models.
\begin{figure}[]
  \centering
  \includegraphics[width=8cm]{IONdensities.eps}
  \caption{\label{IONdensCOMP}
    Mass density profiles for lipids and ions from simulations with different force fields. 
    The lipid densities are divided with 100 to make them visible with the used y-axis scale.
  }
\todo{This figure is in progress, however, the main point can be seen. \\
- I think that we should use number densities instead of mass densities to make the difference between Na and Ca distributions more clear. 
Now the difference comparison is little bit distracted due to the different mass of the atoms. \\
- The lipid density from MacRog simulation is missing. Maybe Matti Javanainen could provide this?
If the data will be uploaded to Zenodo, it also trivial to calculate it with script. \\
- CONCENTRATION CALCULATION ISSUE FIXED, SEE http://nmrlipids.blogspot.com/2015/02/the-first-draft-of-ion-lipid.html?showComment=1423750867868\#c3435589938183508901. \\
The systems are here now with quite different concentrations so comparison is not straigthforward. One option would be to run simulations of all
systems with 150mM which would be close to physiological concentrations, and show the density distributions from those here. \\
- I think that CHARMM36 simulation with CaCl$_2$ should be done \\
- Figure editing will be finished (centering of the graphs, labels, etc.) will be done once the above issues have been sorted out.
 }
\end{figure}
It is evident from the figure that the Na$^+$ density peaks at the lipid bilayer interface are higher (indicating stronger ion binding) for the models
experiencing also larger changes in order parameters with NaCl concentration shown in Fig.~\ref{ordPnacl}. 
The both Na$^+$ binding affinity and order parameter changes are smallest, thus closest to the experiments, for the Orange model, and then incease 
in order CHARMM36, MacRog, and Berger having the highest binding affinity and order parameter changes.
%Simultaneously, it is evident from Figs.~\ref{ordPnacl} and~\ref{ordPions} that the order parameter decrease due to addition
%of NaCl is more pronounced in Berger and MacRog models compared to the CHARMM36 and Orange. 
Thus, the correlation between decreasing choline order parameters
and amount of penetrating cations (i.e. the electrometer concept) is also seen in simulations, independently of the quantitative quality of the used simulation model.
